Efficient management and analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr

### 2017-10-03 cabios-template2/bigsnpr-final.tex
try again

%% LyX 1.3 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english, 12pt]{article}
\usepackage{times}
%\usepackage{algorithm2e}
\usepackage{url}
\usepackage{bbm}
\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\usepackage{geometry}
\geometry{verbose,letterpaper,tmargin=2.5cm,bmargin=2.5cm,lmargin=2.5cm,rmargin=2.5cm}
\usepackage{rotating}
\usepackage{color}
\usepackage{graphicx}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{setspace}
\usepackage{lineno}
\usepackage{hyperref}
\usepackage{bbm}

\linenumbers
\doublespacing
%\usepackage[authoryear]{natbib}
\usepackage{natbib} \bibpunct{(}{)}{;}{author-year}{}{,}

%Pour les rajouts
\usepackage{color}
\definecolor{trustcolor}{rgb}{0,0,1}

\usepackage{adjustbox}
\usepackage{multirow}
\usepackage{graphicx}
\graphicspath{{../figures/}}

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% Bold symbol macro for standard LaTeX users
%\newcommand{\boldsymbol}[1]{\mbox{\boldmath $#1$}}

%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}

\usepackage{babel}
\makeatother

\begin{document}

\title{Efficient management and analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr}
\author{Florian Priv?\,$^{\text{ 1,}*}$, Hugues Aschard\,$^{\text{2,3}}$ and Michael G.B. Blum\,$^{\text{1,}*}$}

\date{~ }
\maketitle

\noindent$^{\text{\sf 1}}$Universit? Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
\noindent$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Int?grative (C3BI), Institut Pasteur, Paris, France \\
\noindent$^{\text{\sf 3}}$Department of Epidemiology, Harvard T.H. Chan School of Public Health, Boston, Massachusetts, USA.

\noindent$^\ast$To whom correspondence should be addressed.

\newpage
\abstract{\textbf{Motivation:} Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals. This increase in data size is a major challenge severely slowing down genomic analyses. Specialized software for every part of the analysis pipeline have been developed to handle large genomic data. However, combining all these software into a single data analysis pipeline might be technically difficult. \\
\textbf{Results:} Here we present two R packages, bigstatsr and bigsnpr, allowing for management and analysis of large scale genomic data to be performed within a single comprehensive framework. To address large data size, the packages use memory-mapping for accessing data matrices stored on disk instead of in RAM. To perform data pre-processing and data analysis, the packages integrate most of the tools that are commonly used, either through transparent system calls to existing software, or through updated or improved implementation of existing methods. In particular, the packages implement a fast derivation of Principal Component Analysis, functions to remove SNPs in Linkage Disequilibrium, and algorithms to learn Polygenic Risk Scores on millions of SNPs. We illustrate applications of the two R packages by analysing a case-control genomic dataset for the celiac disease, performing an association study and computing Polygenic Risk Scores. Finally, we demonstrate the scalability of the R packages by analyzing a simulated genome-wide dataset including 500,000 individuals and 1 million markers on a single desktop computer. \\
\textbf{Availability:} \href{https://privefl.github.io/bigstatsr/}{https://privefl.github.io/bigstatsr/} \& \href{https://privefl.github.io/bigsnpr/}{https://privefl.github.io/bigsnpr/}\\
\textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

\newpage
\section{Introduction}

@@

@@

\section{Methods}

@@

@@

@@

@@

Any test statistic that is based on counts could be easily implemented because we provide fast counting summaries. Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcome \cite[]{Zheng2012}. 
We also implement statistical tests based on linear and logistic regressions. For the linear regression, for each SNP $j$, a t-test is performed on $\beta^{(j)}$ where $$\hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,$$ and $K$ is the number of principal components and $L$ is the number of other covariates (such as the age and gender). Similarly, for the logistic regression, for each SNP $j$, a Z-test is performed on $\beta^{(j)}$ where $$\log{\frac{\hat{p}}{1-\hat{p}}} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,$$ and $\hat{p} = \mathbb{P}(Y = 1)$ and $Y$ denotes the binary phenotype.

@@

@@

@@

@@

@@

\newpage
\begin{figure}[!h]
\centerline{\includegraphics[width=\textwidth]{conversion+QC.pdf}}
\caption{Conversion and Quality Control preprocessing functions available in package bigsnpr via system calls to PLINK.}\label{fig:qc}
\end{figure}

\newpage
\begin{figure}[!h]
\centerline{\includegraphics[width=\textwidth]{imputation.pdf}}
\caption{Imputation and reading functions available in package bigsnpr.}\label{fig:impute}
\end{figure}

\newpage
\begin{figure}[!h]
\centerline{\includegraphics[width=\textwidth]{svd.pdf}}
\caption{Functions available in packages bigstatsr and bigsnpr for the computation of a partial Singular Value Decomposition of a genotype array, with 3 different methods for thinning SNPs.}\label{fig:svd}
\end{figure}

\newpage
\begin{figure}[!h]
\centerline{\includegraphics[width=\textwidth]{celiac-pca}}
\caption{Principal Components of the celiac cohort genotype matrix produced by package bigstatsr.}\label{fig:pca}
\end{figure}

\newpage
\begin{figure}[!h]
\centerline{\includegraphics[width=\textwidth]{celiac-gwas-cut}}
\caption{Manhattan plot of the celiac disease cohort produced by package bigsnpr. Some SNPs in chromosome 6 have p-values smaller than the $10^{-30}$ threshold used for vizualisation purposes.}\label{fig:gwas}
\end{figure}

\newpage
\begin{table}[!h]
\centering
\begin{adjustbox}{max width=\textwidth}
\begin{tabular}{|c|c|c|}
\hline
\multirow{2}{*}{Operation} &   \multicolumn{2}{|c|}{Execution times (in seconds)} \\
 \cline{2-3}
 & PLINK and FastPCA & bigstatsr and bigsnpr \\
\hline
Reading PLINK files & n/a & 5 / 20 \\
Pruning & 4 / 4 & 14 / 52 \\
Computing 10 PCs & 306 / 315 & 58 / 180 \\
GWAS (binary phenotype) & 339 / 293 & 301 / 861 \\
\hline
Total & 649 / 612 & 378 / 1113 \\
\hline
\end{tabular} 
\end{adjustbox}
\caption{Execution times with bigstatsr and bigsnpr compared to PLINK and FastPCA for making a GWAS for the Celiac dataset. The first execution time is with a desktop computer (6 cores used and 64GB of RAM) and the second one is with a laptop computer (2 cores used and 8GB of RAM).} 
\label{tab:bench-gwas}
\end{table}

\newpage
\begin{table}[!h]
\centering
\begin{adjustbox}{max width=\textwidth}
\begin{tabular}{|c|c|c|}
\hline
\multirow{2}{*}{Operation} &   \multicolumn{2}{|c|}{Execution times (in seconds)} \\
 \cline{2-3}
 & PLINK & bigstatsr and bigsnpr \\
\hline
GWAS (binary phenotype) & 232 / 239 & 178 / 650  \\
Clumping & 49 / 58 & 10 / 35 \\
PRS & 9 / 10 & 2 / 3 \\
\hline
Total & 290 / 307 & 190 / 688 \\
\hline
\end{tabular} 
\end{adjustbox}
\caption{Execution times with bigstatsr and bigsnpr compared to PLINK and FastPCA for making a PRS for a training set of 80\% of the Celiac dataset. The first execution time is with a desktop computer (6 cores used and 64GB of RAM) and the second one is with a laptop computer (2 cores used and 8GB of RAM).}
\label{tab:bench-prs}
\end{table}

\newpage
\begin{figure}[!h]
\centerline{\includegraphics[width=\textwidth]{benchmark-pca}}
\caption{Benchmark comparisons between randomized Partial Singular Value Decomposition available in FlashPCA2, FastPCA (fast mode of SmartPCA/EIGENSOFT) and package bigstatsr. It shows the computation time in minutes as a function of the number of samples. The first 10 principal components have been computed based on the 93,083 SNPs which remained after thinning.}\label{fig:bench-pca}
\end{figure}

\begin{figure}[!h]
\centerline{\includegraphics[width=\textwidth]{precision-pca}}
\caption{Precision comparisons between randomized Partial Singular Value Decomposition available in FlashPCA2, FastPCA (fast mode of SmartPCA/EIGENSOFT) and package bigstatsr. It shows the squared correlation between approximated PCs and ``true'' PCs (given by the slow mode of SmartPCA) of the Celiac dataset (whose individuals have been repeated 1, 5 and 10 times).}\label{fig:prec-pca}
\end{figure}

\newpage
\section*{Acknowledgements}

Authors acknowledge Grenoble Alpes Data Institute, supported by the French National Research Agency under the ``Investissements d?avenir'' program (ANR-15-IDEX-02) and the LabEx PERSYVAL-Lab (ANR-11-LABX-0025-01).

@@

\newpage
\section*{Supplementary Data}

\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{table}{0}

\subsection{Long-range LD regions}

\begin{table}[!h]
\centering
\begin{tabular}{lccc}
  \hline
 & Chromosome & Start (Mb) & Stop (Mb) \\ 
  \hline
1 &  2 & 134.7 (134.5) & 137.3 (138) \\ 
  2 &  6 & 27.5 (25.5) & 33.1 (33.5) \\ 
  3 &  8 & 6.6 (8) & 13.2 (12) \\ 
   \hline
\end{tabular}
\caption{Regions found by snp\_autoSVD for the POPRES dataset. Numbers in parentheses correspond to regions referenced in \cite{Price2008}.}
\label{tab:lrldr-popres}
\end{table}

% latex table generated in R 3.3.3 by xtable 1.8-2 package
% Tue Jun 13 11:29:13 2017
\begin{table}[!h]
\centering
\begin{tabular}{l|cccccccccc}
  \hline
 & PC1 & PC2 & PC3 & PC4 & PC5 & PC6 & PC7 & PC8 & PC9 & PC10 \\
  \hline
PC1 & 100.0 & -0.1 & -0.0 & 0.1 & -0.1 & 0.0 & 0.0 & 0.0 & -0.0 & -0.0 \\
  PC2 & 0.1 & 100.0 & -0.0 & 0.1 & -0.0 & -0.0 & -0.0 & 0.2 & -0.1 & -0.0 \\
  PC3 & 0.0 & -0.0 & 99.9 & 0.9 & 0.1 & -0.1 & -0.3 & 0.2 & 0.4 & 0.1 \\
  PC4 & -0.1 & -0.1 & -0.9 & 99.7 & -1.0 & 0.7 & 0.6 & 0.2 & 0.3 & 0.9 \\
  PC5 & 0.1 & 0.0 & -0.1 & 1.1 & 99.3 & 1.3 & -0.8 & 1.3 & -4.2 & -2.4 \\
  PC6 & -0.0 & 0.0 & 0.1 & -0.7 & -1.0 & 97.7 & -3.5 & 6.1 & 7.9 & -6.2 \\
  PC7 & -0.0 & -0.1 & 0.2 & -0.3 & -1.7 & 0.3 & 58.3 & 73.2 & -25.9 & 9.1 \\
  PC8 & 0.1 & -0.1 & -0.3 & 0.4 & -0.5 & -5.3 & -73.5 & 59.5 & 15.8 & 13.2 \\
  PC9 & 0.0 & 0.1 & -0.4 & -0.8 & 5.0 & -7.6 & 27.8 & 11.0 & 91.9 & 9.0 \\
  PC10 & 0.1 & 0.0 & 0.0 & -0.9 & 1.6 & 10.2 & 3.9 & -19.6 & -6.3 & 89.2 \\
   \hline
\end{tabular}
\caption{Correlation between scores of PCA for the POPRES dataset  when automatically removing long-range LD regions and when removing them based on a predefined table.}
\label{tab:pc-popres}
\end{table}

% latex table generated in R 3.3.3 by xtable 1.8-2 package
% Tue Jul 18 15:59:51 2017
\begin{table}[!h]
\centering
\begin{tabular}{lccc}
  \hline
 & Chromosome & Start (Mb) & Stop (Mb) \\ 
  \hline
1 &  2 & 134.4 (134.5) & 138.1 (138) \\ 
  2 &  6 & 23.8 (25.5) & 35.8 (33.5) \\ 
  3 &  8 & 6.3 (8) & 13.5 (12) \\ 
  4 &  3 & 163.1 (n/a) & 164.9 (n/a) \\ 
  5 & 14 & 46.6 (n/a) & 47.5 (n/a) \\ 
   \hline
\end{tabular}
\caption{Regions found by snp\_autoSVD for the celiac dataset. Numbers in parentheses correspond to regions referenced in \cite{Price2008}.}
\label{tab:lrldr-celiac}
\end{table}

% latex table generated in R 3.3.3 by xtable 1.8-2 package
% Tue Jun 13 11:25:56 2017
\begin{table}[!h]
\centering
\begin{tabular}{l|cccccccccc}
  \hline
 & PC1 & PC2 & PC3 & PC4 & PC5 & PC6 & PC7 & PC8 & PC9 & PC10 \\
  \hline
PC1 & 100.0 & -0.1 & -0.1 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 \\
  PC2 & 0.1 & 100.0 & 0.0 & 0.0 & -0.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 \\
  PC3 & 0.1 & -0.0 & 99.9 & 0.2 & -0.0 & 0.1 & 0.1 & 0.1 & 0.0 & -0.1 \\
  PC4 & -0.0 & -0.0 & -0.3 & 99.9 & -0.1 & 0.1 & -0.1 & 0.0 & 0.1 & 0.1 \\
  PC5 & 0.0 & 0.0 & 0.0 & 0.1 & 99.7 & 0.9 & -0.3 & 0.1 & -0.8 & -0.6 \\
  PC6 & -0.0 & 0.0 & -0.1 & -0.2 & -0.8 & 99.6 & 0.5 & -0.5 & -0.2 & -0.4 \\
  PC7 & -0.0 & 0.0 & -0.1 & 0.0 & 0.5 & -0.4 & 98.9 & 3.1 & 0.7 & 1.6 \\
  PC8 & 0.0 & 0.0 & -0.2 & -0.0 & -0.2 & 0.5 & -3.2 & 98.4 & -4.5 & -1.5 \\
  PC9 & -0.0 & -0.0 & -0.0 & 0.0 & 0.6 & 0.1 & -0.7 & 4.6 & 96.9 & -10.7 \\
  PC10 & -0.0 & -0.0 & 0.1 & -0.1 & 0.3 & 0.1 & -1.2 & 1.5 & 8.6 & 92.7 \\
   \hline
\end{tabular}
\caption{Correlation between scores of PCA for the Celiac dataset  when automatically removing long-range LD regions and when removing them based on a predefined table.}
\label{tab:pc-celiac}
\end{table}

\begin{figure}[!h]
\centerline{\includegraphics[width=\textwidth]{scores}}
\caption{PC4 and PC5 of the celiac disease dataset. Left panel, PC scores obtained without removing any long range LD region (only clumping at $R^2 > 0.2$). Individuals are coloured according to their genotype at the SNP that has the highest loading for PC4. Right panel, PC scores obtained with the automatic detection and removal of long-range LD regions. Individuals are coloured according to their population of origin.}\label{fig:scores}
\end{figure}

\begin{figure}[!h]
\centerline{\includegraphics[width=\textwidth]{loadings}}
\caption{Loadings of first 6 PCs of the celiac disease dataset plotted as hexbins (2-D histogram with hexagonal cells). On the left, without removing any long range LD region (only clumping at $R^2 > 0.2$). On the right, with the automatic detection and removal of long-range LD regions.}\label{fig:loadings}
\end{figure}

\begin{figure}[!h]
\centerline{\includegraphics[width=\textwidth]{gini}}
\caption{Boxplots of 1000 bootstrapped Gini coefficients (measure of statistical dispersion) of squared loadings without removing any long range LD region (only clumping at $R^2 > 0.2$) and with the automatic detection and removal of long-range LD regions. The dashed line corresponds to the theoretical value for gaussian loadings.}\label{fig:gini}
\end{figure}

\newpage
\subsection{Imputation}

\begin{figure}[!h]
\centerline{\includegraphics[width=\textwidth]{hist-maf}}
\caption{Histogram of the minor allele frequencies of the POPRES dataset used for comparing imputation methods.}\label{fig:maf}
\end{figure}

\begin{figure}[!h]
\centerline{\includegraphics[width=\textwidth]{hist-NA}}
\caption{Histogram of the number of missing values by SNP. These numbers were generated using a Beta-binomial distribution.}\label{fig:NA}
\end{figure}

\begin{figure}[!h]
\centerline{\includegraphics[width=\textwidth]{error-impute}}
\caption{Number of imputation errors vs the estimated number of imputation errors by SNP. For each SNP with missing data, the number of imputation errors corresponds to the number of individuals for which imputation is incorrect. The estimated number of errors is a quantity that is returned when imputing with snp\_fastimpute, which is based on XGBoost \cite[]{Chen2016}.}\label{fig:error-impute}
\end{figure}

>>>>

\documentclass{bioinfo}
\copyrightyear{2015} \pubyear{2015}

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Manuscript Category}

\usepackage{adjustbox}
\usepackage{multirow}
%\textsc{\usepackage{graphicx}
%\graphicspath{{../figures/}}}
\usepackage{xr}
\externaldocument{bigsnpr-final-supp}

\begin{document}
\firstpage{1}

\subtitle{Subject Section}

\title[R packages for analyzing genome-wide data]{Efficient management and analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr}
\author[Sample \textit{et~al}.]{Florian Privé\,$^{\text{\sfb 1,}*}$, Hugues Aschard\,$^{\text{\sfb 2,\sfb 3}}$ and Michael G.B. Blum\,$^{\text{\sfb 1,}*}$}
\address{$^{\text{\sf 1}}$Université Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Intégrative (C3BI), Institut Pasteur, Paris, France and \\
$^{\text{\sf 3}}$Department of Epidemiology, Harvard T.H. Chan School of Public Health, Boston, Massachusetts, USA.}

\corresp{$^\ast$To whom correspondence should be addressed.}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

\abstract{\textbf{Motivation:} Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals. This increase in data size is a major challenge severely slowing down genomic analyses. Specialized software for every part of the analysis pipeline have been developed to handle large genomic data. However, combining all these software into a single data analysis pipeline might be technically difficult. \\
\textbf{Results:} Here we present two R packages, bigstatsr and bigsnpr, allowing for management and analysis of large scale genomic data to be performed within a single comprehensive framework. To address large data size, the packages use memory-mapping for accessing data matrices stored on disk instead of in RAM. To perform data pre-processing and data analysis, the packages integrate most of the tools that are commonly used, either through transparent system calls to existing software, or through updated or improved implementation of existing methods. In particular, the packages implement a fast derivation of Principal Component Analysis, functions to remove SNPs in Linkage Disequilibrium, and algorithms to learn Polygenic Risk Scores on millions of SNPs. We illustrate applications of the two R packages by analysing a case-control genomic dataset for the celiac disease, performing an association study and computing Polygenic Risk Scores. Finally, we demonstrate the scalability of the R packages by analyzing a simulated genome-wide dataset including 500,000 individuals and 1 million markers on a single desktop computer. \\
\textbf{Availability:} \href{https://privefl.github.io/bigstatsr/}{https://privefl.github.io/bigstatsr/} \& \href{https://privefl.github.io/bigsnpr/}{https://privefl.github.io/bigsnpr/}\\
\textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

\maketitle

\section{Introduction}

@@

%\enlargethispage{12pt}

@@

\begin{methods}
\section{Methods}

@@

\begin{figure}[!tpb]
\centerline{\includegraphics[width=160pt]{conversion-QC.pdf}}
\caption{Conversion and Quality Control preprocessing functions available in package bigsnpr via system calls to PLINK.}\label{fig:qc}
\end{figure}

@@

\begin{figure}[!tpb]
\centerline{\includegraphics[width=235pt]{imputation.pdf}}
\caption{Imputation and reading functions available in package bigsnpr.}\label{fig:impute}
\end{figure}

@@

\begin{figure}[!tpb]
\centerline{\includegraphics[width=235pt]{svd.pdf}}
\caption{Functions available in packages bigstatsr and bigsnpr for the computation of a partial Singular Value Decomposition of a genotype array, with 3 different methods for thinning SNPs.}\label{fig:svd}
\end{figure}

@@

Any test statistic that is based on counts could be easily implemented because we provide fast counting summaries. Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcome \cite[]{Zheng2012}. 
We also implement statistical tests based on linear and logistic regressions. For the linear regression, for each SNP $j$, a t-test is performed on $\beta^{(j)}$ where
\begin{multline}
  \hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K \\ + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,
\end{multline}
and $K$ is the number of principal components and $L$ is the number of other covariates (such as the age and gender). Similarly, for the logistic regression, for each SNP $j$, a Z-test is performed on $\beta^{(j)}$ where
\begin{multline}
  \log{\frac{\hat{p}}{1-\hat{p}}} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K \\ + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,
\end{multline}
and $\hat{p} = \mathbb{P}(Y = 1)$ and $Y$ denotes the binary phenotype.

@@

\end{methods}

@@

\begin{figure}[!tpb]
\centerline{\includegraphics[width=200pt]{celiac-pca}}
\caption{Principal Components of the celiac cohort genotype matrix produced by package bigstatsr.}\label{fig:pca}
\end{figure}

\begin{figure}[!tpb]
\centerline{\includegraphics[width=235pt]{celiac-gwas-cut}}
\caption{Manhattan plot of the celiac disease cohort produced by package bigsnpr. Some SNPs in chromosome 6 have p-values smaller than the $10^{-30}$ threshold used for vizualisation purposes.}\label{fig:gwas}
\end{figure}

@@

\begin{table}[!tpb]
\centering
\begin{adjustbox}{max width=235pt}
\begin{tabular}{|c|c|c|}
\hline
\multirow{2}{*}{Operation} &   \multicolumn{2}{|c|}{Execution times (in seconds)} \\
 \cline{2-3}
 & PLINK and FastPCA & bigstatsr and bigsnpr \\
\hline
Reading PLINK files & n/a & 5 / 20 \\
Pruning & 4 / 4 & 14 / 52 \\
Computing 10 PCs & 306 / 315 & 58 / 180 \\
GWAS (binary phenotype) & 339 / 293 & 301 / 861 \\
\hline
Total & 649 / 612 & 378 / 1113 \\
\hline
\end{tabular} 
\end{adjustbox}
\caption{Execution times with bigstatsr and bigsnpr compared to PLINK and FastPCA for making a GWAS for the Celiac dataset. The first execution time is with a desktop computer (6 cores used and 64GB of RAM) and the second one is with a laptop computer (2 cores used and 8GB of RAM).} 
\label{tab:bench-gwas}
\end{table}

\begin{table}[!tpb]
\centering
\begin{adjustbox}{max width=235pt}
\begin{tabular}{|c|c|c|}
\hline
\multirow{2}{*}{Operation} &   \multicolumn{2}{|c|}{Execution times (in seconds)} \\
 \cline{2-3}
 & PLINK & bigstatsr and bigsnpr \\
\hline
GWAS (binary phenotype) & 232 / 239 & 178 / 650  \\
Clumping & 49 / 58 & 10 / 35 \\
PRS & 9 / 10 & 2 / 3 \\
\hline
Total & 290 / 307 & 190 / 688 \\
\hline
\end{tabular} 
\end{adjustbox}
\caption{Execution times with bigstatsr and bigsnpr compared to PLINK and FastPCA for making a PRS for a training set of 80\% of the Celiac dataset. The first execution time is with a desktop computer (6 cores used and 64GB of RAM) and the second one is with a laptop computer (2 cores used and 8GB of RAM).}
\label{tab:bench-prs}
\end{table}

@@

\begin{figure}[!tpb]
\centerline{\includegraphics[width=200pt]{benchmark-pca.png}}
\caption{Benchmark comparisons between randomized Partial Singular Value Decomposition available in FlashPCA2, FastPCA (fast mode of SmartPCA/EIGENSOFT) and package bigstatsr. It shows the computation time in minutes as a function of the number of samples. The first 10 principal components have been computed based on the 93,083 SNPs which remained after thinning.}\label{fig:bench-pca}
\end{figure}

\begin{figure}[!tpb]
\centerline{\includegraphics[width=200pt]{precision-pca.png}}
\caption{Precision comparisons between randomized Partial Singular Value Decomposition available in FlashPCA2, FastPCA (fast mode of SmartPCA/EIGENSOFT) and package bigstatsr. It shows the squared correlation between approximated PCs and ``true'' PCs (given by the slow mode of SmartPCA) of the Celiac dataset (whose individuals have been repeated 1, 5 and 10 times).}\label{fig:prec-pca}
\end{figure}

@@

\section*{Acknowledgements}

Authors acknowledge Grenoble Alpes Data Institute, supported by the French National Research Agency under the ``Investissements d'avenir'' program (ANR-15-IDEX-02) and the LabEx PERSYVAL-Lab (ANR-11-LABX-0025-01).

@@
